### 2017-07-02 opsem.tex
Sec.3 text part 1.

% Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
% \txnimp, a core language that we will use to formalize the intuitions
% presented in the previous section. Natural numbers, (shared) variables
% and arithmetic expressions constitute the syntactic class of
% expressions ($e$).  Commands ($c$) include $\cskip$, assignment
% statements, transaction (\C{txn}) lexical blocks, and their sequential
% and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
% over transaction identifiers. When it is evident we are referring to a
% transaction, we use the number $i$ instead of $T_i$ for identification
% (\eg in $\C{txn}\langle i \rangle$). Like variables, transaction
% identifiers are globally accessible. For notational convenience, we
% let $t$ range over both expressions and commands.

% We define a small-step operational semantics for this language in
% terms of an abstract machine that generates an execution trace ($\E$).
% The first component of the trace is a set ($\A$) of \emph{effects},
% where an effect ($\eta$) is a tuple of (a). the identifier ($T_i$) of
% the transaction that generated this effect, (b). the unique
% identifier ($j$) of this effect, (c). the operation ($\op$)
% documented by this effect, which can be either a read (\C{RD(X)}), a
% write (\C{WR(X)}), or a transaction commit (\C{COMMIT}), and (d). the
% value, if any, associated with the operation (\eg, value read or value
% written). Functions $\txn$, $\id$, $\oper$, and $\rval$ project each
% of the four aforementioned individual components of the tuple.  In
% every step of the evaluation, the machine reduces a \txnimp\ term by
% executing a read, write or commit operation, generating an effect, and
% extending the trace. Since effects include transaction identifiers,
% the semantics distinguishes between terms ($t$) of different
% transactions. For example, $\txnbox{t}_i$ denotes a term $t$ inside a
% transaction $T_i$.  Evaluation contexts are also appropriately marked.
% For example, $\ectx_i$ denotes the evaluation context for a term
% inside $T_i$. The other component of an execution trace is a
% visibility relation ($\visZ$) that establishes a visibility property
% between effects among different transactions. The intent and mechanics
% of $\visZ$ is described in the sequel.

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), Integer and boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\{\bar{f}=\bar{e}\}$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$), and
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Certain terms that only appear at run-time are also
present in $c$. These include a \C{TXN} block tagged with sets ($\stl$
and $\stg$) of records that assume special meaning in operational
semantics, and a \C{FOREACH} loop that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that even the surface-level \C{FOREACH} shown here is a
little different from the one used in previous sections; its
higher-order argument has two arguments, $y$ and $z$, which are
invoked (during the reduction) with the set of already-iterated items,
and the current item, respectively. This form of \C{FOREACH} lends
itself to inductive reasoning, facilitating inductive proofs
(Sec.~\ref{sec:reasoning}).

% also We let $T_i$ for $i \in \mathbb{N}$
% range over transaction identifiers. When it is evident we are
% referring to a transaction, we use the number $i$ instead of $T_i$ for
% identification (\eg in $\C{txn}\langle i \rangle$). Like variables,
% transaction identifiers are globally accessible. For notational
% convenience, we let $t$ range over both expressions and commands.

We define a small-step operational semantics for this language in
terms of an abstract machine that executes a command, and updates
either a transaction-local database ($\stl$), or the global database
($\stg$). Database ($\stg$) is a modeled as a set of records of a
pre-defined type, i.e., they all belong to a single table.
Generalization to multiple tables is straightforward. Records in
$\stg$ are uniquely identifiable through their $\idf$ field, which is
auto-generated and does not belong to the surface language, i.e.,
$\C{id}\notin f$. For a set $S$ of records, we define $\dom(S)$ as the
set of unique ids of all the records in $S$. Thus $|\dom(\stg)| =
|\stg|$. During its execution, a transaction may write to multiple
records in $\stg$. Atomicity dictates that such writes shouldn't be
available in $\stg$ until the transaction commits. We therefore
associate each transaction with a local database ($\stl$) that stores
\emph{only} the uncommitted records\footnote{While SQL's \C{UPDATE}
admits writes at the granularity of record fields, databases, in
reality, enforce record-level locking, allowing us to think of
``uncommitted writes'' as ``uncommitted records''. }. Uncommitted
records include deleted records, which are identified by a hidden
$\delf$ field set to \C{true}. When the transaction commits, its local
database is atomically \emph{flushed} to the global database,
committing the uncommitted records. The flush operation ($\gg$) is defined as
following:
\begin{smathpar}
\begin{array}{c}
\forall r.~ r \in (\stl\gg\stg) ~\Leftrightarrow~ 
  (r.\idf \notin \dom(\stl) \conj r \in \stg)
\disj (r \in \stl \conj \neg r.\delf) 
\end{array}
\end{smathpar}
Let $\stg' = \stl\gg\stg$. A record $r$ belongs to $\stg'$ iff it
belongs to $\stg$ and not been updated in $\stl$, i.e., $r.\idf \notin
\dom(\stl)$, or it belongs to $\stl$, i.e., it is either a new record,
or an updated version of an old record, but the updatation is not a
deletion ($\neg r.\delf$). Thus, flush defines the result of
atomically applying transaction's local writes to the global database.
Besides the commit, flush also helps a transaction read its own
writes. Intuitively, the result of a read operation inside a
transaction must be computed on the database resulting from flushing
the current local state ($\stl$) to the global state ($\stg$). The
abstract machine of Fig.~\ref{fig:txnimp}, however, doesn't let a
transaction read its own writes. This reduces verbosity and simplifies
semantics, without losing any generality; substituting $\stl\gg\stg$
for $\stg$ at select places in reduction rules recovers the real

Small-step semantics is stratified into a transaction-local reduction
relation, and a top-level reduction relation. Transaction-local
relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
small-step reduction for a command inside a transaction, when the
database state is $\stg$; the command $c$ reduces to $c'$, while
updating the transaction-local database $\stl$ to $\stl'$. The
definition assumes a meta function $\eval$ that evaluates expressions
with no free variables to values. The reduction relation for SQL
statements is defined straightforwardly.  \C{INSERT} adds a new record
to $\stl$ after adding a unique identifier (more discussion on
uniqueness later). \C{DELETE} finds the records in $\stg$ that match
the search criteria defined by its boolean function argument, and adds
the records to $\stl$ after marking them for deletion. \C{SELECT}
bounds the name introduced by \C{LET} to the set of records from
$\stg$ that match the search criteria, and then executes the bound
command $c$. \C{UPDATE} uses its first function argument to compute
the updated version of the records that match the search criteria
defined by its second function argument. Updated records are added to
$\stl$.

The reduction of \C{FOREACH} starts by first converting it to its
run-time form that also keeps track of the iterated items ($s_1$),
besides the yet-to-be-iterated items ($s_2$). Initially, $s_1$ is
empty. As the elements are iterated, they are removed from $s_2$ and
added to $s_1$ . Iteration involves invoking the higher-order function
with $s_1$ and the current element $x$ (note: $\uplus$ in $\{x\}
\uplus s_2$ denotes a disjoint union). The reduction ends when the
$s_2$ becomes empty. The reduction rules for conditionals, \C{LET}
binders, and sequences are straightforward, hence ommitted.

The top-level reduction relation
